

\chapter*{ABSTRACT}
\phantomsection
\addcontentsline{toc}{chapter}{ABSTRACT}

\noindent The Uptime Tracker is a web-based monitoring system designed to track website availability and performance metrics in real-time. In today's digital landscape, website downtime can significantly impact productivity, service delivery, and user trust. This project addresses the need for automated monitoring of college websites and web services by implementing a centralized tracking system similar to status.tcioe.edu.np and downdetector.com.

\noindent The system is built using Bun runtime with Next.js for the frontend and Elysia.js for the backend API, providing fast and efficient full-stack performance. PostgreSQL serves as the database for reliable data storage. The platform features user authentication, website registration and ownership management, automated ping checks at 5-minute intervals, and comprehensive analytics dashboards. It monitors HTTP/HTTPS availability, measures response times, and calculates hourly and daily average uptime statistics. The application supports both public and private monitoring modes, allowing website owners to manage visibility of their site's status.

\noindent The Uptime Tracker provides real-time status information, historical data visualization, and performance analytics that enable administrators to quickly identify and respond to downtime incidents. This transparency improves user experience and supports data-driven decisions for service reliability improvements. The project documentation is prepared using LaTeX for professional formatting.

\vspace{1em}
\noindent\textit{\textbf{Keywords}}---Website Monitoring, Uptime Tracking, Bun, Next.js, Elysia.js, PostgreSQL, Real-time Status, Performance Analytics, Web Application.